Dissecting the relationship between protein structure and sequence variation

### 2014-11-21 manuscript/manuscript_cordiv.tex

%%  address and funding acknowledgement data:
\def\affiliation{Department of Physics \\
The University of Texas at Austin, TX 78712, USA; amir@physics.utexas.edu}
%\def\support{\footnote{Work supported by the Office of Science,Kingdom of the Two Sicilies, under contract OSS--32456.}}

@@

\vfill
\Title{Sequence Divergence as the main modulator of structure-sequence correlations}
\vfill
\Author{Amir Shahmoradi, Claus Wilke, Eleisha Jackson, ...}%\support}
\Address{\affiliation}
\vfill
\begin{Abstract}
    Over the past decade several independent works have shown that some site-specific structural properties of proteins are capable of predicting the site-specific evolutionary sequence variations. The strength and significance of these structure-evolution relations, however, appear to vary widely among different proteins, with absolute correlation strengths ranging from $0.1$ to $0.8$. In search for potential determinants of the observed variation in sequence-structure correlations, here we present the results of a comprehensive study of interrelations among more than $200$ structural and evolutionary characteristics in a dataset of 209 monomeric enzymes. We identify sequence divergence as the main determinant of the strengths of virtually all structure-evolution relationships, explaining $\sim10-30\%$ of observed variation in correlation strengths. In addition to sequence divergence, we identify several protein structural properties that are weakly but significantly coupled with the strength of sequence-structure relations. In particular, proteins with more homogeneous back-bone hydrogen bond energies, large fractions of helical secondary structures and low fraction of beta sheets tend to exhibit the strongest sequence-structure relations.
\end{Abstract}
\vfill
\vfill
\end{titlepage}
\def\thefootnote{\fnsymbol{footnote}}
\setcounter{footnote}{0}
%

@@

        Throughout this work, we have carried out a comprehensive analysis in search for the main determinants of the strength of sequence--structure correlations -- some of which are newly reported and discussed in this work. Examples of sequence--structure relations include the correlations of sequence entropy ({\it seqent}) and measures of evolutionary rates (such as {\it r4sJC} used in this work) with measures of residue Contact Number (e.g., wcnSC), Relative Solvent Accessibility (RSA), \ddg entropy ({\it ddgent}). In addition, we have derived new site-specific properties based on Voronoi Tessellation of protein 3D structures that are comparable to or better than several previously known structural properties in explaining site specific sequence entropy or evolutionary rates (e.g., Figure \ref{fig:seqent_structure_cors}). Prime examples include Voronoi cell volume ({\it vvolume}), surface area ({\it varea}) and Voronoi cell sphericity defined as,

@@

        in which $V$ \& $A$ represent {\it vvolume} \& {\it varea} respectively. We have also shown that site-specific structural properties -- such as Weighted Contact Number, Bfactor and Voronoi Cell properties -- that are calculated from the average coordinates of side chain atoms, have the best explanatory powers for the sequence variability measures such as {\it seqent} and {\it r4sJC}. Compared to the common choice of back bone {\it CA} atomic coordinates, site-specific properties averaged over side chain atoms can outperform in predicting sequence evolutionary rates $0.12$ in terms of Spearman correlation strength.

@@

        \begin{figure}[tbh]
            \begin{center}
            \includegraphics[width=6.9in]{cordiv_similarities_r4sJC.pdf}
            \end{center}
            \caption{A comparison of the strength of Spearman correlation of sequence evolutionary rates (r4sJC) with {\it side chain} Weighted Contact Number vs. correlations of other structural properties with evolutionary rates. Detailed description of the structural properties is given in Section \label{sec:mam}. The red lines in each plot represent equality line. It is evident from all plots that for any given protein in dataset, the correlation strength of one structural property is a good proxy measure of the correlation strength of any other structural property with sequence variability measures. For brevity, correlations of structure-rate4site are not shown here but are available online, also in supplementary material.}
            \label{fig:seqent_structure_cors}
        \end{figure}

@@

        \begin{figure}[tbh]
            \begin{center}
            \includegraphics[width=6.9in]{/best_vorvol/selected_variables/boxplot_vorvol_all_in_one.pdf}
            \end{center}
            \caption{Should be likely modified to volume-area-length-sphericity measures instead.}
            \label{fig:best_voronoi}
        \end{figure}

        \begin{figure}[tbh]
            \centering
            \includegraphics[width=7.2in]{/main_modulators_cormat.pdf} \\
            \caption{{\bf The Spearman correlation matrix of the strongest sequence--structure correlations and the prominent determinants of the strengths of the corresponding relations}. The variables on the diagonal elements of the matrix from top to bottom represent respectively, the four strongest sequence--structure relations: r4sJC--wcnSC, r4sJC--varea, seqent--wcnSC, seqent--varea, followed by protein properties that modulate the strength of these relations: the correlation between r4sJC--seqent ({\it r.r4sJC.seqent}), variance of r4sJC ({\it var.r4sJC}), variance of sequence entropy ({\it var.seqent}), variance of back-bone hydrogen bond energies ({\it var.hbe}), and the fraction of amino acids in helical \& $\beta-$sheet secondary structures in the proteins ({\it mn.helix} \& {\it mn.betas} respectively).}
            \label{fig:mainmods}
        \end{figure}

@@

        \begin{figure*}[tbh]
            \centering
            \begin{tabular}{cc}
            \includegraphics[width=3.4in]{/validation_sdseqent.pdf} & \includegraphics[width=3.4in]{/validation_sdhbe.pdf} \\
            \end{tabular}
            \caption{{\bf Sequence--structure correlation strength versus sequence divergence.} The plot illustrates the relationship between the strength of a representative sequence--structure correlation ({\it seqent--scnSC}) and the sequence divergence as measured by the variance of protein sequence entropy. The black circles represent $209$ proteins used in this work. For comparison and validation, the red circles represent data for $9$ viral proteins taken from {shahmoradi2014xx}.
            \label{fig:validation}
        \end{figure*}

>>>>

%%  address and funding acknowledgement data:
\def\affiliation{Department of Physics \\
The University of Texas at Austin, TX 78712, USA; amir@physics.utexas.edu}
%\def\support{\footnote{Work supported by the BEACON-NSF center for the study of evolution in action, under contract xxxxxx.}}

@@

\vfill
%\Title{Sequence Divergence as the main modulator of structure-sequence correlations}
\Title{Dissecting the relationship between protein structure and sequence variation}
\vfill
\Author{Amir Shahmoradi, Eleisha Jackson, Claus Wilke}%\support}
\Address{\affiliation}
\vfill
\begin{Abstract}
    Over the past decade several independent works have shown that some structural properties of proteins are capable of predicting protein evolution. The strength and significance of these structure-sequence relations, however, appear to vary widely among different proteins, with absolute correlation strengths ranging from $0.1$ to $0.8$. Here we present the results from a comprehensive search for the potential biophysical and structural determinants of protein evolution by studying more than $200$ structural and evolutionary properties in a dataset of $209$ monomeric enzymes. We discuss the main protein characteristics responsible for the general patterns of protein evolution, and identify sequence divergence as the main determinant of the strengths of virtually all structure-evolution relationships, explaining $\sim10-30\%$ of observed variation in sequence-structure relations. In addition to sequence divergence, we identify several protein structural properties that are moderately but significantly coupled with the strength of sequence-structure relations. In particular, proteins with more homogeneous back-bone hydrogen bond energies, large fractions of helical secondary structures and low fraction of beta sheets tend to have the strongest sequence-structure relation.

    %Over the past decade several independent works have shown that some site-specific structural properties of proteins are capable of predicting the site-specific evolutionary sequence variations. The strength and significance of these structure-evolution relations, however, appear to vary widely among different proteins, with absolute correlation strengths ranging from $0.1$ to $0.8$. In search for potential determinants of the observed variation in sequence-structure correlations, here we present the results of a comprehensive study of interrelations among more than $200$ structural and evolutionary characteristics in a dataset of 209 monomeric enzymes. We identify sequence divergence as the main determinant of the strengths of virtually all structure-evolution relationships, explaining $\sim10-30\%$ of observed variation in correlation strengths. In addition to sequence divergence, we identify several protein structural properties that are weakly but significantly coupled with the strength of sequence-structure relations. In particular, proteins with more homogeneous back-bone hydrogen bond energies, large fractions of helical secondary structures and low fraction of beta sheets tend to exhibit the strongest sequence-structure relations.
\end{Abstract}
\vfill
\vfill
\end{titlepage}
\def\thefootnote{\fnsymbol{footnote}}
\setcounter{footnote}{0}
%

@@

        Throughout this work, we have carried out a comprehensive analysis in search for the main determinants of the strength of sequence--structure correlations -- some of which are newly reported and discussed in this work. Examples of sequence--structure relations include the correlations of sequence entropy ({\it seqent}) and measures of evolutionary rates (such as {\it r4sJC} used in this work) with measures of residue Contact Number (e.g., {\it wcnSC}), Relative Solvent Accessibility (RSA), \ddg entropy ({\it ddgent}). In addition, we have derived new site-specific properties, based on Voronoi Tessellation of protein 3D structures, that are comparable to or better than several previously known structural properties in explaining site-specific sequence entropy or evolutionary rates (e.g., Figure \ref{fig:seqent_structure_cors}). Prime examples include Voronoi cell volume ({\it vvolume}), surface area ({\it varea}) and Voronoi cell sphericity defined as,

@@

        in which $V$ \& $A$ represent {\it vvolume} \& {\it varea} respectively. We have also shown that site-specific structural properties -- such as Weighted Contact Number, Bfactor and Voronoi Cell properties -- that are calculated from the average coordinates of side chain atoms, have the best explanatory powers for the sequence variability measures such as {\it seqent} and {\it r4sJC}. Compared to the common choice of back bone {\it CA} atomic coordinates, site-specific properties averaged over side chain atoms can outperform in predicting sequence evolutionary rates by as much as $0.12$ in terms of Spearman correlation strength.

        In search for the determinants of the strength of sequence-structure relations, we compiled a set of more than $200$ protein properties for a dataset of $209$ monomeric enzymes. By employing several independent parametric and non-parametric statistics, such as Spearman rank test, regularized regression and Principal Regression methods, we identify sequence divergence as the dominant factor in the strength of sequence-structure correlations, capable of explaining $10-30\%$ of the observed correlation strengths alone, in both the original and rank-transformed data.

@@

%\clearpage
%\newpage

%\section*{Figures}

    \begin{figure}[tbh]
            \begin{center}
            %\includegraphics[width=6.9in]{cordiv_similarities_r4sJC.pdf}
            \includegraphics[width=6.9in]{cordiv_similarities_seqent.pdf}
            \end{center}
            \caption{A comparison of the strength of Spearman correlation of sequence evolutionary rates (r4sJC) with {\it side chain} Weighted Contact Number vs. correlations of other structural properties with evolutionary rates. Detailed description of the structural properties is given in Section \label{sec:mam}. The red lines in each plot represent equality line. It is evident from all plots that for any given protein in dataset, the correlation strength of one structural property is a good proxy measure of the correlation strength of any other structural property with sequence variability measures. For brevity, correlations of structure-rate4site are not shown here but are available online, also in supplementary material.}
            \label{fig:seqent_structure_cors}
    \end{figure}

    \begin{figure}[tbh]
        \begin{center}
        \includegraphics[width=6.9in]{/best_wcn/selected_variables/boxplot_wcn_two_in_one.pdf}
        \end{center}
        \caption{A comparison of the predictive power of different measures of Weighted Contact Number (WCN) about different structure or sequence properties for 209 proteins in dataset. In each plot, the Spearman correlation strengths of a given structure or sequence property on the vertical axis with different measures of WCN on the horizontal axis are compared against each other. The capital letters in the variable names on the horizontal axis denote the set of atomic coordinates used to calculate WCN. The variable $wcnSC$ denotes WCN measure based on the average coordinates over all heavy atoms in the side chain and the variable $wcnAA$ denotes WCN measure based on the average coordinate over all heavy atoms in the side chain and backbone of the amino acid.}
        \label{fig:best_wcn}
    \end{figure}

    \begin{figure}[tbh]
        \centering
        \includegraphics[width=3.6in]{/main_modulators_cormat_r4s_wcn.pdf} \\
        \caption{{\bf The Spearman correlation matrix of the strongest sequence--structure correlations and the prominent determinants of the strengths of the corresponding relations}. The variables on the diagonal elements of the matrix from top to bottom represent respectively, the four strongest sequence--structure relations: r4sJC--wcnSC, r4sJC--varea, seqent--wcnSC, seqent--varea, followed by protein properties that modulate the strength of these relations: the correlation between r4sJC--seqent ({\it r.r4sJC.seqent}), variance of r4sJC ({\it var.r4sJC}), variance of sequence entropy ({\it var.seqent}), variance of back-bone hydrogen bond energies ({\it var.hbe}), and the fraction of amino acids in helical \& $\beta-$sheet secondary structures in the proteins ({\it mn.helix} \& {\it mn.betas} respectively).}
        \label{fig:mainmods}
    \end{figure}

    \begin{figure*}[tbh]
        \centering
        %\begin{tabular}{cc}
        \includegraphics[width=3.4in]{/validation_r_r4s_wcn_var_seqent.pdf}
        %\end{tabular}
        \caption{{\bf Sequence--structure correlation strength versus sequence divergence.} The plot illustrates the relationship between the strength of a representative sequence--structure correlation ({\it seqent--scnSC}) and the sequence divergence as measured by the variance of protein sequence entropy. The black circles represent $209$ proteins used in this work. For comparison and validation, the red circles represent data for $9$ viral proteins taken from {shahmoradi2014xx}}.
        \label{fig:validation}
    \end{figure*}

%\newpage

%\section*{Supplementary Figures}

    %\centerline{\includegraphics[width=6.5in]{PC_screen_entropy_CSrmsf.pdf}}
    %\noindent \textbf{Fig. S1} Principal Component (PC) Regression of sequence entropy against the structural variables, including CS RMSF. {\bf (A)} Variance in entropy explained by each principal component. For most proteins, PC1 and either PC2 or PC3 show the strongest correlations with sequence entropy. Significant correlations ($P<0.05$) are shown as filled symbols, and insignificant correlations ($P\geq0.05$) are shown as open symbols. {\bf (B)} and {\bf (C)} Composition of the three leading components. Red arrows represent the loadings of each of the structural variables on the principal components; black dots represent the amino acid sites in the PC coordinate system.
    %\customlabel{fig:cor_entropy_PC_screen_CSrmsf}{S1}

@@

    \begin{figure}[tbh]
        \begin{center}
        \includegraphics[width=6.9in]{/best_varea/selected_variables/boxplot_varea_all_in_one.pdf}
        \end{center}
        \caption{Should be likely modified to volume-area-length-sphericity measures instead.}
        \label{fig:best_voronoi}
    \end{figure}

@@

    \begin{figure*}[tbh]
        \centering
        %\begin{tabular}{cc}
        \includegraphics[width=3.4in]{/validation_r_seqent_wcn_var_seqent.pdf}
        %\end{tabular}
        \caption{{\bf Sequence--structure correlation strength versus sequence divergence.} The plot illustrates the relationship between the strength of a representative sequence--structure correlation ({\it seqent--scnSC}) and the sequence divergence as measured by the variance of protein sequence entropy. The black circles represent $209$ proteins used in this work. For comparison and validation, the red circles represent data for $9$ viral proteins taken from {shahmoradi2014xx}}.
        \label{fig:validation}
    \end{figure*}
